On the effects of partying on next-day sports performance

### 2017-05-30 writing/draft.tex
Major updates through results section. Added party_discrete results.

\usepackage{hyperref}
\usepackage{color,soul}

@@

\paragraph{Logic for using spread data}
In order to assert a causal relation between cities with active nightlife 
and subsequent day sports performance, we need to control for salient
features such as the ability of players on either team in the match-up,
whether any players are injured, the momentum of success each team is carrying
in the season, etc. To that end, we utilize spread data. The benefit of
using spread data is that it controls for a lot, but the downside is that
we don't know exactly what it controls for.

For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically over-valued
teams which you are more likely to play after a party city,
we could see a spurious effect.

\hl{Careful} That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reliable. So, if we do find an effect in our variable of interest,
we can be confident in our conclusion that there is a causal effect and not some
other systematic bias.

@@

\paragraph{The problem of an unknown travel schedule} In NBA, we have the difficulty of
not observing when teams travel. 
In MLB, each match-up is actually a series of games.

\subsection{Data}
\subsubsection{Lines data} We obtained data from \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{Covers.com}, which records the outcome of each game alongside the line set by a betting house.

\subsubsection{BLS data} We also obtained data from
\href{https://www.bls.gov/data/}{Bureau of Labor Statistics} which records the number of establishments by business type at the county-quarter level. As a 
proxy for how much night-life there is in a city, we look toward the 
number of sound recording studios, musical groups, and music publishers there 
are for a particular county-year: we simply calculate the average number of establishments across all three music categories listed above.\footnote{We caveat that since the Toronto Raptors are located in Canada, we don't have BLS data for this team.}

We then merge this in with our lines data, taking care to do so according to each season-team's last game location; we also lag our BLS data by one year.
To this data, we've created features such as 
the number of days since last game, the last game location, the travel distance, and
average age\footnote{We use $\log \textrm{\# days old}$, at the time the game is played.}. Our data range from 2010-2011 season through present, which includes part of the 2016-2017 season.

\subsubsection{Minutes data}
We also gathered minutes data from \href{http://www.espn.com/nba/scoreboard/_/date/}{ESPN}.

The minutes data also contain information on the number of free throws, field goals, and
three pointers made and attempted by each player in each game.

\subsubsection{Loctaion and Travel Distance Data} is obtained
by API.

\subsection{Method: Defining a continuous measure of night life}

We have a free parameter: the time-window which we believe the hangover effect
to last. Suppose we only look toward games which have been played in a party
city at most 24 hours ago. We take care to ignore
teams playing in their own cities.

\paragraph{Discrete indicator}
If a team plays in Los Angeles or New York within 24 hours of their current game,
we indicate one, else zero.

\paragraph{Continuous measure}
If the last game was played at most 36 hours ago, use log average number
of musicians in last game location, else zero.

\subsection{Model} We use a logistic regression to estimate the likelihood of meeting
the spread as a function of whether the team partied as well as controlling for other confounders
listed above.

\section{Results}
\subsection{Using spread data}
\begin{center}
\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]  \\[-1.8ex] & outcome == "W" \\  \hline \\[-1.8ex]   party & $-$0.052$^{**}$ (0.022) \\    lag.chg.pos & 0.0001 (0.002) \\    travel.dist & $-$0.00002 (0.00004) \\    ew & $-$0.035 (0.047) \\    nhours.lgame & 0.001 (0.001) \\    I(hour(date)) & 0.006 (0.012) \\    travel.dist:ew & 0.00004 (0.00005) \\    Constant & $-$0.144 (0.348) \\   \textit{N} & 9,979 \\  Log Likelihood & $-$6,912.426 \\  Akaike Inf. Crit. & 13,840.850 \\  \hline  \hline \\[-1.8ex]  \textit{Notes:} & \multicolumn{1}{r}{$^{***}$Significant at the 1 percent level.} \\   & \multicolumn{1}{r}{$^{**}$Significant at the 5 percent level.} \\   & \multicolumn{1}{r}{$^{*}$Significant at the 10 percent level.} \\  \end{tabular} 
\end{center}

\subsection{Using performance statistics}
We additionally looked at actual performance instead of relying on spread data. 
This allows us to avoid using the black-box of spread data, and also pin-point
more exactly what elements of the game are in fact effected by partying.
This motivates looking at number of points allowed, number of points scored.

We also considered rebounds, fouls, and even injuries. 
\begin{center}
    \begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]  \\[-1.8ex] & demeaned.reb.per.min \\  \hline \\[-1.8ex]   party & $-$0.001$^{*}$ (0.0005) \\    lag.chg.pos & 0.00003 (0.0001) \\    travel.dist & 0.00000$^{**}$ (0.00000) \\    nhours.lgame & 0.00002 (0.00002) \\    Constant & $-$0.003$^{**}$ (0.002) \\   \textit{N} & 44,190 \\  R$^{2}$ & 0.0003 \\  Adjusted R$^{2}$ & 0.0003 \\  Residual Std. Error & 0.105 (df = 44185) \\  F Statistic & 3.857$^{***}$ (df = 4; 44185) \\  \hline  \hline \\[-1.8ex]  \textit{Notes:} & \multicolumn{1}{r}{$^{***}$Significant at the 1 percent level.} \\   & \multicolumn{1}{r}{$^{**}$Significant at the 5 percent level.} \\   & \multicolumn{1}{r}{$^{*}$Significant at the 10 percent level.} \\  \end{tabular}  
\end{center}

\subsection{Robustness check}
We verify that this phenomena persists in the MLB. In MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and party. Additionally, we interacted weekend.
One of the major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

\begin{center}
    \begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]  \\[-1.8ex] & I(team.score \textgreater  opponent.score) \\  \hline \\[-1.8ex]   party & $-$0.034$^{*}$ (0.019) \\    odds & 2.602$^{***}$ (0.162) \\    I(team == location) & 0.028 (0.033) \\    ndays.lgame & 0.012 (0.019) \\    travel.dist & 0.00000 (0.00003) \\    weekend & 0.042 (0.036) \\    Constant & $-$1.333$^{***}$ (0.080) \\   \textit{N} & 25,215 \\  Log Likelihood & $-$17,274.460 \\  Akaike Inf. Crit. & 34,562.920 \\  \hline  \hline \\[-1.8ex]  \textit{Notes:} & \multicolumn{1}{r}{$^{***}$Significant at the 1 percent level.} \\   & \multicolumn{1}{r}{$^{**}$Significant at the 5 percent level.} \\   & \multicolumn{1}{r}{$^{*}$Significant at the 10 percent level.} \\ \end{tabular}  
\end{center}

\subsubsection{Betting performance}
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred during the season.
    We first train a model on historical data then use said model to predict on ``current year''
    data, ploting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet if our model instructs us of odds sufficiently different from what
    the betting house suggests. We account for the bookmaker's cut. E.g. in the early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}

@@

\section{This should be the first section of your document}

Your text goes here. Start with the first section or paragraph of your
article. Do not set use either headers or footers and do not set any running
heads or change any page numbers.

The title page along with headers and footers will be inserted by the EdiKit
system. Use the ``revise manuscript'' link to enter this information in the
EdiKit system.

Use the standard LaTeX commands to set the text of the article. The dgjournal
package works with any of the standard LaTeX document classes - article, report
or book. Other document classes should work but compliance with De Gruyter
requirements is not assured. For more information on LaTeX, see
\cite{lamport,mittelbach,oetiker}. \cite{mittelbach} is highly recommended.

Use the standard LaTeX sectioning commands for your headings.

\subsection{A second order heading}

Some text under the subheading. Paragraphs that follow heads are not
indented.

Math should also be set in Times. Use the mathptmx package if you do not have
any of the commercially available fonts that are compatible with Times.
\begin{equation}
    y^{(n)} = \sum_{i=0}^{n-1} a_i(x) y^{(i)} + r(x) 
\end{equation}

All environments provided by the standard LaTeX document classes are
unchanged. Vertical spaces within lists have been altered to comply with De Gruyter
requirements.
\begin{enumerate}
\item This is the first item within the list. Some more text here in order to
  display the alignment.
\item Another item in the list.
\item Yet another item in the list.
\end{enumerate}

Here is an example of a Figure. It's the same as in standard LaTeX.

\begin{figure}[!h]
%% Use the graphics package to insert figures
%% \includegraphics{figure.eps}
% Use \centering to center the table
\centering
%% A small box in place of a figure
\framebox{%
  \begin{minipage}{10pc}
    \begin{center}
      \vspace{1cm}\par
      A figure\par
      \vspace{1cm}
    \end{center}
  \end{minipage}}
\caption{Insert your caption here. If you wish to label your figure for
  cross-referencing, use a label either within the caption or after it.}
\label{fig1}
\end{figure}

An example of a table follows. This is also the same as in standard LaTeX.

\begin{table}[!h]
% Use \centering to center the table
\centering
\caption{Insert your table caption here. If you wish to label the table for
  cross-referencing, use a label either within the caption or after it.}
\begin{tabular}{llll}
\hline
Symbol        & LaTeX Command      & Symbol      & LaTeX Command \\
\hline
$\alpha$      & \verb+\alpha+      & $\zeta$     & \verb+\zeta+ \\
$\beta$       & \verb+\beta+       & $\eta$      & \verb+\eta+ \\
$\gamma$      & \verb+\gamma+      & $\theta$    & \verb+\theta+ \\
$\delta$      & \verb+\delta+      & $\vartheta$ & \verb+\vartheta+ \\
$\epsilon$    & \verb+\epsilon+    & $\iota$     & \verb+\iota+ \\
$\varepsilon$ & \verb+\varepsilon+ & $\kappa$    & \verb+\kappa+ \\
\hline
\end{tabular}
\end{table}

Use the \verb+thebibliography+ environment for the references.  BibTeX users may
use the provided BibTeX style file DeGruyter.bst.

>>>>

\usepackage{hyperref}
\usepackage{color,soul}
\usepackage{amsmath}

@@

\paragraph{Logic for using spread data}
In order to assert a causal relation between cities with active nightlife 
and subsequent day sports performance, we need to control for salient
features such as the ability of players on either team in the match-up,
whether any players are injured, the momentum of success each team is carrying
in the season, etc.\cite{anderson} To that end, we utilize spread data. The benefit of
using spread data is that it controls for a lot, but the downside is that
we don't know exactly what it controls for.

For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically over-valued
teams which you are more likely to play after a party city,
we could see a spurious effect. 
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reliable. So, if we do find an effect in our variable of interest,
we can be confident in our conclusion that there is a causal effect and not some
other systematic bias.

@@

\paragraph{The problem of an unknown travel schedule} In NBA, we have the difficulty of not observing when teams travel. E.g. if a particular team  plays
in Los Angeles on Friday and another game in New York the following day, we don't
know whether the team traveled on Friday evening or Saturday morning. Heuristics for
travel schedules can be made, but it is hard to obtain reliable team-travel data.
In MLB, on the other hand, each match-up is actually a series of games. E.g.
if the Dodgers play the Giants in San Francisco, they will play several games spread
across several days, each in the same city; in these cases, we can infer that the team
is in fact spending the night in said city.

\subsection{Data}
\paragraph{Lines data} We first look toward 
\href{http://www.basketball-reference.com/leagues/}{basketball-reference.com}
for a listing of game-days by season. With this listing, we queried \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{covers.com},
for all games occurring on a particular game, and for each we observe
outcome of each game alongside the line set by a betting house. We obtain
data for seasons 2010-11 through 2016-17.

\paragraph{BLS data} We also obtained data from
\href{https://www.bls.gov/data/}{Bureau of Labor Statistics} which records the number of establishments by business type at the county-quarter level. As a 
proxy for how much night-life there is in a city, we look toward the 
number of sound recording studios, musical groups, and music publishers there 
are for a particular county-year: we simply calculate the average number of establishments across all three music categories listed above. We caveat that since the Toronto Raptors are located in Canada, we don't have BLS data for this team.

We then merge this in with our lines data, taking care to do so according to each season-team's last game location such that the feature captures the proper
party effect. In addition, we lag our BLS data by one year for two reasons:
realistically, player expectations of which cities have the most active night-life may not
update in real-time and also doing so allows us to create a betting model, i.e.
we don't rely on any features we can't obtain before making a game-time prediction. 
To this data, we've created features such as 
the number of days since last game, the last game location, and the travel distance. 
Our data range from 2010-2011 season through present, which includes part of the 2016-2017 season.

\paragraph{Minutes data}
We also gathered minutes data from \href{http://www.espn.com/nba/scoreboard/_/date/}{ESPN}.
The minutes data also contain information on the number of free throws, field goals, and three pointers made and attempted by each player in each game.

\paragraph{Location and Travel Distance Data} We are also interested
to account for the effects of jet-lag. For each team, we queried
a maps search engine to obtain a longitude and latitude listing.\cite{ggmap}
For each pairwise set of coordinates, we calculated distance between the two
points.\cite{sp} To account for jet-lag, we first computed the final direction 
(bearing) when arriving at the 
second point after starting from the first and following the shortest path on a sphere 
(following a great circle).\cite{lallensack}

\subsection{Method: Defining what it means for a city to have nightlife}
We have a free parameter: the time-window which we believe the hangover effect
to last. Suppose we only look toward games which have been played in a party
city at most 24 hours ago. We take care to ignore
teams playing in their own cities, i.e. if a team's last game was played in at home 
we say that they were not exposed to any nightlife.

\paragraph{Discrete indicator}
Our first sanity check was to define an indicator variable for when a team plays in 
Los Angeles or New York within 24 hours of their current game, we indicate one, else zero. Anecdotally, these are mentioned as hot-spots for
nightlife.

\paragraph{Continuous measure}
We also defined a continuous measure of nightlife:
if the last game was played at most 24 hours ago, use the log of the average number
of musicians in last game location, else emit zero. Intuitively, this captures
how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

\subsection{Model} We use a logistic regression to estimate the likelihood of meeting
the spread as a function of whether the team partied as well as controlling for other confounders
listed above. Our basic model is as follows.

\[
\texttt{meet spread} \,  \sim \, \substack{\texttt{days since} \\ \texttt{last game}} \, + \, \texttt{jetlag} \, + \, \substack{\texttt{home team} \\ \texttt{effect}} \, + \, \texttt{party}
\]

\section{Results}
\subsection{Using spread data}
We find that after accounting for player fatigue incurred during the last game, jetlag and travel fatigue, and recovery time, and home-team effect, that both our discrete indicator and continuous measure of party
have a statistically significant effect on likelihood of meeting the spread.

\subsection{Using performance statistics}
We additionally looked at actual performance instead of relying on spread data. 
This allows us to avoid using the black-box of spread data, and also pin-point
more exactly what elements of the game are in fact effected by partying.
This motivates looking at number of points allowed, number of points scored.
We also considered rebounds, fouls, and even injuries.

In particular, we see that the average number of rebounds obtained per minute
of game-play does decrease as a function of being exposed to a city with active night-life.
The method was to first calculate for each player and each game the number of rebounds obtained
per minute of game-play. Then, for each player and each season we subtract from each observation
the player's average number of rebounds obtained per minute of game-play using the entire season of data. This is an ad-hoc approach that cannot be used for betting.

\subsection{Robustness check}
We verify that this phenomena persists in Major League Baseball. In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and our party variable. Additionally, we interacted our party variable with weekend.
One of the major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

% Table created by stargazer v.5.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu % Date and time: Mon, May 29, 2017 - 11:56:17 PM
\hspace{-105pt}
\begin{tabular}{@{\extracolsep{5pt}}lcccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{4}{c}{\textit{Dependent variable:}} \\  \cline{2-5}  \\[-1.8ex] & \multicolumn{2}{c}{NBA: Meet Spread} & demeaned reb/min & MLB: Win Game \\  \\[-1.8ex] & \multicolumn{2}{c}{\textit{logistic}} & \textit{OLS} & \textit{logistic} \\  \\[-1.8ex] & (1) & (2) & (3) & (4)\\  \hline \\[-1.8ex]  party.discrete & $-$0.406$^{***}$ (0.133) &  &  &  \\    party &  & $-$0.053$^{**}$ (0.022) & $-$0.001$^{*}$ (0.0005) & $-$0.035$^{*}$ (0.020) \\    lag.chg.pos & 0.0001 (0.002) & 0.0001 (0.002) & 0.00003 (0.0001) &  \\    odds &  &  &  & 2.602$^{***}$ (0.162) \\    log(travel.dist + 1) & $-$0.017 (0.025) & $-$0.015 (0.025) & 0.001 (0.001) & $-$0.001 (0.005) \\    ew & $-$0.114 (0.230) & $-$0.101 (0.230) &  &  \\    nhours.lgame & 0.001 (0.001) & 0.001 (0.001) & 0.00003 (0.00002) &  \\    I(hour(date)) & 0.007 (0.012) & 0.005 (0.012) &  &  \\    weekend &  &  &  & 0.040 (0.036) \\    I(team == location) & $-$0.011 (0.043) & $-$0.001 (0.043) &  & 0.025 (0.034) \\    log(travel.dist + 1):ew & 0.018 (0.035) & 0.016 (0.035) &  &  \\    ndays.lgame &  &  &  & 0.014 (0.019) \\    Constant & $-$0.077 (0.374) & $-$0.041 (0.374) & $-$0.007$^{*}$ (0.004) & $-$1.331$^{***}$ (0.080) \\   \hline \\[-1.8ex]  Observations & 10,000 & 9,979 & 44,190 & 25,215 \\  R$^{2}$ &  &  & 0.0003 &  \\  Adjusted R$^{2}$ &  &  & 0.0002 &  \\  Log Likelihood & $-$6,925.180 & $-$6,912.606 &  & $-$17,274.450 \\  Akaike Inf. Crit. & 13,868.360 & 13,843.210 &  & 34,562.900 \\  Residual Std. Error &  &  & 0.105 (df = 44185) &  \\  F Statistic &  &  & 3.064$^{**}$ (df = 4; 44185) &  \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

\subsubsection{Betting performance}
We also validate our model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred during the season.
    We first train a model on historical data then use said model to predict on ``current year''
    data, ploting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet if our model instructs us of odds sufficiently different from what
    the betting house suggests. We account for the bookmaker's cut. E.g. in the early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}

\clearpage

@@
